Paper Summary: Uncertainty Visualisation

### 2025-11-30 twopagesummary/summary.qmd
set up summary document

---
title: "The Noisy Work of Uncertainty Visualisation Research: A Review"
author: 
  - name: Harriet Mason
    url: https://harrietmason.netlify.app/
    orcid: 0009-0007-4568-8215
    email:  harriet.m.mason@gmail.com
    affiliation:
    - name: Monash University
      department: Department of Econometrics and Business Statistics
      city: Melbourne
      country: Australia
  - name: Dianne Cook
    url: https://dicook.org
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu
    affiliation:
    - name: Monash University
      department: Department of Econometrics and Business Statistics
      city: Melbourne
      country: Australia
  - name: Sarah Goodwin
    url: https://www.linkedin.com/in/smgoodwin/
    orcid: 0000-0001-8894-8282
    email: Sarah.Goodwin@monash.edu
    affiliation:
    - name: Monash University
      department: Department of Human Centred Computing
      city: Melbourne
      country: Australia
  - name: Emi Tanaka
    url: https://emitanaka.org/
    orcid: 0000-0002-1455-259X
    email: Emi.Tanaka@anu.edu.au
    affiliation: 
    - name: The Australian National University
      department: Biological Data Science Institute
      city: Canberra
      country: Australia
  - name: Susan VanderPlas
    url: https://srvanderplas.github.io
    orcid: 0000-0002-3803-0972
    email: susan.vanderplas@unl.edu
    affiliation:
    - name: University of Nebraska–Lincoln
      department: Statistics Department
      city:  Lincoln
      country: United States
bibliography: paper.bib
abstract: Uncertainty visualisation is quickly becoming a hot topic in information visualisation. Existing reviews in the field take the definition and purpose of an uncertainty visualisation to be self evident which results in a large amout of conflicting information. This conflict largely stems from a conflation between uncertainty visualisations designed for decision making and those designed to prevent false conclusions. We coin the term "signal suppression" to describe a visualisation that is designed for preventing false conclusions, as the approach demands that the signal (i.e. the collective take away of the estimates) is suppressed by the noise (i.e. the variance on those estimates). We argue that the current standards in visualisation suggest that uncertainty visualisations designed for decision making should not be considered uncertainty visualisations at all. Therefore, future work should focus on signal suppression. Effective signal suppression requires us to communicate the signal and the noise as a single "validity of signal" variable, and doing so proves to be difficult with current methods. We illustrate current approaches to uncertainty visualisation by showing how they would change the visual apprearance of a choropleth map. These maps allow us to see why some methods succeed at signal suppression, while others fall short. Evaluating visualisations on how well they perform signal suppression also proves to be difficult, as it involves measuring the effect of noise, a variable we typically try to ignore. We suggest authors use qualitative studies or compare uncertainty visualisations to the relevant hypothesis tests. 
date: last-modified
toc: false
number-sections: true
latex-clean: true
format:
  jasa-pdf:
    keep-tex: true  
    journal:
      blinded: false
  jasa-html: default
fig-valign: bottom
cap-location: bottom
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| message: false
#| warning: false

# load Libraries
library(tidyverse)
# devtools::install_github("lydialucchesi/Vizumap")
library(Vizumap)
library(RColorBrewer)
library(scales)
library(sf)
library(ggrepel)
# devtools::install_github("UrbanInstitute/urbnmapr")
library(urbnmapr)
library(flextable)
library(colorspace)
#library(rgeos)
```

## Introduction

From entertainment choices to news articles to insurance plans, the modern citizen is so inundated with information in every aspect of their life it can be overwhelming. 
In the face of this overflow of information, tools that effectively reduce piles of information to simple and clear ideas become more valuable. 
That is, we need tools that can sort the signal from the noise.

Among these summary tools, information visualisations are one of the most powerful as they allow for quick and memorable communication that allow us to identify quirks in our data that we didn't know to look for.
Datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurus; @datasaurpkg] show a case where visual statistics highlight elements of the data that are invisible to the typical summary statistics. 
Something as simple as sketching a distribution before recalling statistics or making predictions can greatly increase the accuracy of those measures [@Hullman2018; @Goldstein2014].

"Uncertainty visualisation" is a relatively new field in research. 
Early mentions of uncertainty visualisation start to appear in the late 1980s [@Ibrekk1987], with geospatial information visualisation literature from the early 1990s declaring this to be essential aspect of any information display [@MacEachren1992; @Carr1992]. 
@fig-ibrekk depicts an example of the uncertainty visualisations discussed in these early papers. 
Despite kicking off the field, these papers did not define uncertainty visualisation.
This has led to a lack of consensus on what it means for a graphic to visualise uncertainty, an issue we will return to later.
Therefore, while the field is considered to be quite new, many of the graphics used for uncertainty visualisation have been around for much longer. 
For example, box plots and histograms display variation which becomes synonymous with uncertainty when we are using them to depict the variation of an estimate. 
Today, there are an abundance of publications on the topic which makes it timely to construct a review of the field.
That is, now that there is an overwhelming amount of information, it is valuable to distill it into simple facts. 
In fact, there have already been several reviews published but a central piece of discussion is missing.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-ibrekk
#| fig-cap: "A replication of the uncertainty visualisations shown by @Ibrekk1987 in one of the earliest uncertainty visualisation experiments. Several visualisation methods that are now unpopular (such as the pie chart) are used throughout this paper."
#| fig-subcap: 
#|   - "Picture 1"
#|   - "Picture 2"
#|   - "Picture 3"
#|   - "Picture 4"
#|   - "Picture 5"
#|   - "Picture 6"
#|   - "Picture 7"
#|   - "Picture 8"
#|   - "Picture 9"
#| layout-ncol: 3

# Generate data
set.seed(1)
x=rnorm(1000, 8, 4)
ib_data <- tibble(x=ifelse(x<0, -x, x))

# Picture 1
p1 <- ib_data |>
  summarise(avg = mean(x),
          conf_95a = quantile(x, probs=c(0.025)),
          conf_95b = quantile(x, probs=c(0.975))) |>
  ggplot(aes(y="NA")) +
  geom_point(aes(x=avg)) +
  geom_errorbar(aes(xmin = conf_95a, xmax = conf_95b), width = 0.1) +
  scale_x_continuous(name = "INCHES OF SNOW",
                     breaks=seq(0,19),
                     labels= ggplot2:::interleave(as.character(c(seq(0,18, 2), 19)), rep("", 11))[c(0:19, 21)],
                     limits=c(0,19)) +
  theme_classic() +
  theme(axis.line.y=element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        aspect.ratio=1/10)

# Picture 2
p2 <- ib_data |>
  mutate(x = ifelse(x>18, 18, x),
         binx = cut(x, breaks=seq(0,18,2))) |>
  group_by(binx) |>
  summarise(n = n()) |>
  mutate(Probability = n / sum(n)) |>
  ggplot(aes(x=binx, y=Probability)) +
  geom_col(fill="black", colour="white") +
  scale_x_discrete(name = "INCHES OF SNOW",
                   labels= paste0(seq(0,16,2), sep = "-", seq(2,18,2))) +
  scale_y_continuous(breaks = seq(0.00, 0.25, 0.05)) +
  theme_classic() + 
  theme(aspect.ratio=0.33)

# Picture 3
p3 <- ib_data |>
  mutate(x = ifelse(x>18, 18, x),
         binx = cut(x, 
                     breaks=seq(0,18,2), 
                     labels= paste0(seq(0,16,2), sep = "-", seq(2,18,2)))) |>
  group_by(binx) |>
  summarise(n = n()) |>
  mutate(Probability = n / sum(n),
         csum = rev(cumsum(rev(Probability))), 
         pos = Probability/2 + lead(csum, 1),
         pos = if_else(is.na(pos), Probability/2, pos)) |>
  ggplot(aes(x="", y=Probability, fill=binx)) +
  geom_bar(stat="identity", width=1) +
  geom_text_repel(aes(y = pos, label = paste0(round(Probability*100), sep="", "%")),
                   size = 3, nudge_x = 0.6, show.legend = FALSE, segment.color = 'transparent') +
  #geom_label(aes(label = paste0(round(Probability*100), sep="", "%")),
  #          position = position_stack(vjust = 0.5)) + 
  scale_fill_grey() +
  coord_polar("y", start=0) +
  labs(fill = "INCHES OF SNOW") + 
  theme_void() + 
  theme(aspect.ratio=1)

# Picture 4
p4 <- ib_data |>
  ggplot(aes(x=x)) +
  geom_density() + 
  scale_x_continuous(name = "INCHES OF SNOW",
                     breaks = seq(0,20,2),
                     labels= paste0(seq(0,20,2))) +
  scale_y_continuous(name = "Probability density",
                     breaks = seq(0.00, 0.20, 0.02)) +
  theme_classic() + 
  theme(aspect.ratio=0.33)

# Picture 5
p5 <- ib_data |>
  ggplot(aes(y="", x=x)) +
  geom_violin() + 
  scale_x_continuous(name = "INCHES OF SNOW",
                     breaks = seq(0,20,2),
                     labels= paste0(seq(0,20,2)),
                     limits=c(0,20)) +
  theme_classic() + 
  theme(axis.line.y=element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        aspect.ratio=0.4)

# Picture 6
set.seed(1)
x=rnorm(5000, 8, 4)
ib_data2 <- tibble(x=ifelse(x<0, -x, x)) |>
  mutate(x=ifelse(x>=18, 18-rexp(5000,rate=0), x))
p6 <- ib_data2 |>
  ggplot(aes(y="", x=x)) +
  geom_jitter(size=0.05) + 
  scale_x_continuous(name = "INCHES OF SNOW",
                     breaks = seq(0,20,2),
                     labels= paste0(seq(0,20,2)),
                     limits=c(0,20)) +
  theme_classic() + 
  theme(axis.line.y=element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        aspect.ratio=0.1)

# Picture 7
p7 <- ib_data2 |>
  arrange(x) |>
  mutate(group = rep(1:50, each=100))|>
  group_by(group) |>
  summarise(x = max(x, na.rm=TRUE)) |>
  add_row(group=c(0,51), x = c(0,20)) |>
  ggplot(aes(x=x)) +
  geom_linerange(ymin = 0.1, ymax = 1) + 
  geom_linerange(y=1, xmin = -0.03, xmax = 20.03)+ 
  geom_linerange(y=0.1, xmin = -0.03, xmax = 20.03)+ 
  scale_x_continuous(name = "INCHES OF SNOW",
                     breaks = seq(0,20,2),
                     labels= paste0(seq(0,20,2)),
                     limits=c(0,20)) +
  scale_y_continuous(limits=c(0,1)) + 
  theme_classic() + 
  theme(axis.line.y=element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        aspect.ratio=0.1)

# Picture 8
p8 <- ib_data |>
  reframe(x = quantile(x, probs=c(0.25, 0.50, 0.75)))|>
  add_row(x = c(0,20)) |>
  arrange(x) |>
  mutate(quantile = c("min", "q1", "med", "q3", "max")) |>
  pivot_wider(names_from = quantile, values_from = x) |>
  ggplot(aes(y="")) +
  #geom_point(aes(x=med)) +
  geom_errorbar(aes(y="", xmin = min, xmax = max), width = 0.2) +
  geom_crossbar(aes(y="", x=med, xmin = q1, xmax = q3), width = 0.5) +
  scale_x_continuous(name = "INCHES OF SNOW",
                     breaks=seq(0,20),
                     labels= ggplot2:::interleave(as.character(c(seq(0,20, 2))), rep("", 11))[1:21],
                     limits=c(0,20)) +
  theme_classic() +
  theme(axis.line.y=element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        aspect.ratio=1/10)

# Picture 9
p9 <- ib_data |>
  ggplot(aes(x)) +
  stat_ecdf(geom = "step") +
  scale_x_continuous(name = "INCHES OF SNOW",
                     breaks=seq(0,20),
                     labels= ggplot2:::interleave(as.character(c(seq(0,20, 2))), rep("", 11))[1:21],
                     limits=c(0,20)) +
  scale_y_continuous(name = "Cumulative probability",
                     breaks=seq(0,1,0.1),
                     labels= seq(0,1,0.1),
                     limits=c(0,1)) +
  theme_classic() +
  theme(aspect.ratio=4/10)

# Display Plots
p1
p2
p3
p4
p5
p6
p7
p8
p9
```

Reviews on uncertainty visualisation rarely offer tried and tested rules for effective uncertainty visualisation, but rather, they comment on the *difficulties* faced when trying to summarise the field. 
@Kinkeldey2014 found most experimental methods to be ad hoc, with no commonly agreed upon methodology, formalisations, or greater goal of describing general principals. 
@Hullman2016 noticed there is a serious noise issue in the data coming from uncertainty visualisation experiments. 
She commented on the prevalence of confounding variables that make it unclear as to what exactly caused a subjects poor performance on a set of particular questions.
Mistakes due to misunderstanding visualisations, misinterpreting questions, and incorrectly applying heuristics are all combined into a single error value.
@Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal best plot for all people and circumstances.
@Griethe2006 did not identify common themes, but instead listed the findings and opinions of a collection of papers. 
@uncertchap2022 summarised several cognitive effects that repeatedly arise in uncertainty visualisation experiments, however these effects were each discussed in isolation as a list of considerations an author might make rather than an overarching theory of rules for effective uncertainty visualisation. 
While these reviews are thorough in scope, none discuss how the existing literature contribute to the broader goal of uncertainty visualisation. 
The problem faced by the literature is easily summarised with a famous quote by Henri Poincaré.

> "Science is built up of facts, as a house is built of stones; but an accumulation of facts is no more a science than a heap of stones is a house." - Henri Poincaré (1905)

That is to say, despite the wealth of reviews, the field of uncertainty visualisation remains a heap of stones. 
There is a mountain of work that identifies common heuristics found in uncertainty visualisations, evaluate competing plot designs, or start a theoretical discussion on a niche aspect of the field.
This is important work that needs to be done, but each of these papers offers up their own bespoke motivation and methodology, with little reference to the uncertainty visualisation papers outside their periphery.
This becomes even more difficult to manage when these studies are in conflict. 
The field is in desperate need of a unifying theory that can tie this swath of research together.
This review attempts to address this issue by offering a novel perspective on the uncertainty visualisation problem.
That is, we are going to use the wealth of established stone to construct the foundations on which we can build a house.

This review is broken into several parts that each reflect a different approach to uncertainty visualisation. 
First, we look at graphics that ignore uncertainty entirely and discuss why uncertainty should be included at all. 
Second, we look at methods that consider uncertainty to be just another variable and discuss the characteristics of uncertainty that make it a unique visualisation problem. 
Third, we look at methods that explicitly combine our estimate and its uncertainty and discuss the limitations of these approaches. 
Fourth, we discuss methods that implicitly include uncertainty by depicting a sample in place of an estimate. 
Finally, we discuss how uncertainty visualisations can be effectively evaluated.
When discussing each of these methods, we will repeatedly return to the *purpose* of uncertainty visualisation and the effectiveness of each approach in fulfilling that purpose.

### Spatial example
There are far too many uncertainty visualisations to exhaustively discuss them all. 
Instead we focus on the changes made to a single plot, the choropleth map. 
Due to the field's origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty. 
Utilising a single example will help isolate the ideas we are trying to convey. 
Therefore, it is important to remember that even though we focus our discussion on the choropleth map, the theoretical approach we outline in this review is useful to all uncertainty visualisations regardless of their application. 
Additionally, our examples focus on incorporating uncertainty through colour manipulation, as that is the key visual channel used in a choropleth map.
However, the methods we discuss go beyond variations in a colour palette.
Even though they are not explicitly shown, visualisations that depict uncertainty using layers such as position or shape, and more complicated graphics that incorporate animation or interactivity are also within the scope of this review.
We will use the choropleth maps as a tool to clearly highlight the costs and benefits of each approach.

@fig-data shows the first six rows and the geographical boundaries of our data set. 
The temperature variable was generated using a sine wave, that is $Temperature_i = 29-2\cdot{|Latitude_i - \sin{2 \cdot Longitude_i}|}$ where the $Longitude$ and $Latitude$ are the longitude and latitude of the county's centroid scaled to a standard normal distribution. 
Each county's variance is independently, randomly sampled from a uniform distribution. 
The low standard errors are drawn from a $U_{[0,1]}$ distribution, while the high standard errors are drawn from a $U_{[1,2]}$ distribution. 
As we are dealing with an average, the sampling distribution would be approximately normal by the central limit theorem, so each county temperature estimate is assumed to come from a $N(Temp_i, SE_{case,i})$. 
This is the data we will be using in our spatial uncertainty examples for the rest of the paper.

```{r}
#| eval: false
#| echo: false
# Get map data: do this once and save
my_map_data <- get_urbn_map("counties", sf = TRUE) |>
  filter(state_name=="Iowa")
save(my_map_data, file="data/iowa_map.rda")
```

```{r}
#| eval: false
#| echo: false
# Get centroid because rgeos is depreciated
load("data/iowa_map.rda")
centroids <- as_tibble(gCentroid(as(my_map_data$geometry, "Spatial"), byid = TRUE))
my_map_data$cent_long <- centroids$x
my_map_data$cent_lat <- centroids$y
save(my_map_data, file="data/my_map_data.rda")
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-data
#| fig-cap: "The first 5 observations of the data used for the spatial uncertainty examples along with the boundaries of each county. The map boundaries are the Iowa county boundaries, however the 'temperature' data is not representative of the average temperature in Iowa. The temperature and standard error represent the average of the daily high temperature and the standard error of that average respectively."
#| fig-subcap: 
#|   - "Data Table"
#|   - "Map Boundaries"
#| layout-nrow: 1
#| cap-location: "bottom"

# get data
load("data/my_map_data.rda")

# seed for sampling
set.seed(1997)

# data dimension for sampling
n <- dim(my_map_data)[1]

# Make palettes
longpal <- rev(sequential_hcl(13, palette = "YlOrRd"))
basecols <- longpal[3:10]
breaks <- 21:29 
breakslong <- 18:32 
names(basecols) <- seq(8)
names(longpal) <- -1:11

my_map_data <- my_map_data |>
  mutate(temp = 29 - 2*abs(scale(cent_lat) - sin(2*(scale(cent_long)))[,1]), # trend
         highvar = runif(n, min=2, max=4), # high variance
         lowvar = runif(n, min=0, max=2), # low variance
         count_id = row_number()) |>
  pivot_longer(cols=highvar:lowvar, names_to = "variance_class", values_to = "variance") |>
  # add bivariate classes to data
  mutate(bitemp = cut(temp, breaks=breaks, labels=seq(8)),
         bivar = cut(variance, breaks=0:4, labels=seq(4)),
         biclass = paste(bitemp, bivar, sep="-"))|>
  mutate(highlight = ifelse(count_id <= 5, TRUE, FALSE))

# Make nice example table
example_table <- my_map_data |>
  mutate(variance = sqrt(variance)) |>
  select(c(count_id, county_name, temp, variance_class, variance)) |>
  as_tibble()|>
  pivot_wider(id_cols=c(count_id, county_name, temp,), 
              names_from = variance_class, 
              values_from = variance) |>
  head(5)|>
  flextable() |>
  set_caption(caption = "Average Daily High Temperatures of Iowa Counties") |>
  add_header_row(colwidths = c(3, 2),
                 values = c("", "Standard Error")) |>
  colformat_double(digits = 2) |>
  set_header_labels(count_id = "ID", 
                    county_name  = "County",
                    temp = "Average Temperature (°C)",
                    highvar = "High",
                    lowvar = "Low") |>
  add_footer_row(values = rep("..."), colwidths = 5) |>
  theme_vanilla() |>
  vline(i=c(1,2), j=3, part="header")|>
  align(align = "left", part = "all") |>
  bg(j = "temp",
    bg = col_numeric(palette = brewer.pal(8, name = "Oranges"),
                     domain = c(21, 30)),
    part = "body"
  ) |>
  bg(j = c("highvar", "lowvar"),
    bg = col_numeric(palette = brewer.pal(8, name = "Greens"),
                     domain = c(0, 3)),
    part = "body"
  )

# make blank map 
example_map <- my_map_data |>
  filter(variance_class=="lowvar") |>
  ggplot() +
  geom_sf(aes(geometry = geometry, fill=highlight)) + 
  scale_fill_manual(values=c("white", "#fbfba2")) +
  geom_text(data=filter(my_map_data, highlight==TRUE), aes(x=cent_long, y=cent_lat, label=count_id), size=3) +
  theme_void() +
  theme(legend.position = 'none')

plot(example_table)

example_map

```

## Ignoring uncertainty
A good place to start might be at a deceptively straight forward question, why should we include uncertainty at all?

### The choropleth map
@fig-choropleth depicts a choropleth map of the counties of Iowa. 
Each of these counties are coloured according to an estimate of average daily temperature that was generated so that the values followed a clear spatial trend. 
The variance of these estimates were simulated such that the trend accounts for most of the variance in the plot in the low variance case (so we should expect the trend to be visible) while in the high variance, there is more variance within each county than between all the counties, so we should expect it to (at least in some capacity) overwhelm the spatial trend. 
Is this aspect of the data and the spatial data it communicates clear in in the map? 
Is the strength of the trend communicated through the visualisation?

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-choropleth
#| fig-cap: "Two choropleth maps that depict the counties of Iowa where each country coloured acording to a simulated average temperature. Both maps depict a spatial trend, where counties closer to the centre of the map are hotter than counties on the edge of the map. In the low variance condition, the trend accounts for most of the variation in the data, in the high variance case, the variance on the temperature estimate accounts for most of the variance. This distinction is not clear in the map as they both appear identical. The high variance condition displays a spatial trend that could simply be spurious, which means the plot is displaying a false conclusion."
#| fig-subcap: 
#|   - "Low Variance Data"
#|   - "High Variance Data"
#|   - "Choropleth Palette"
#| layout-ncol: 3
#| layout-valign: "bottom"
#| cap-location: "bottom"

# Choropleth Map
p1a <- my_map_data |>
  filter(variance_class=="lowvar") |>
  ggplot() +
  geom_sf(aes(fill = bitemp, 
              geometry = geometry), colour=NA) + 
  scale_fill_manual(values = basecols) +
  #scale_fill_gradientn(colours = basecols, 
  #                     values=breaks/limits[2],
  #                     limits=limits) +
  theme_void() + 
  theme(legend.position = "none")

p1b <- p1a %+% filter(my_map_data, variance_class=="highvar")

show_pal <- function (colours, borders = NULL, cex_label = 1, ncol = NULL, myxlab, breaks, textnudge, xlabx, xlaby, tsize=1.2) {
  # Set dimensions of palette
  n <- length(colours)
  ncol <- ncol %||% ceiling(sqrt(length(colours)))
  nrow <- ceiling(n/ncol)
  # make matrix with null values (if not full)
  colours <- c(colours, rep(NA, nrow * ncol - length(colours)))
  colours <- matrix(colours, ncol = ncol, byrow = TRUE)
  # set graphical parameters (?)
  old <- par(pty = "s", mar = c(0, 0, 0, 0))
  on.exit(par(old))
  size <- max(dim(colours))
  plot(c(0, size), c(0, -size), type = "n", xlab = "", ylab = "", 
      axes = FALSE)
  rect(col(colours) - 1, -row(colours) + 1, col(colours), -row(colours), 
       col = colours, border = borders)
  text(c(0,col(colours)) + textnudge, -c(1,row(colours))-0.25, breaks, 
              cex = 1, col = "black")
  text(xlabx, xlaby, myxlab ,cex = tsize, col = "black")
}

p1a
p1b
show_pal(basecols, ncol=8, borders=NA, myxlab = "Temperature", breaks = 21:29, textnudge = c(0.2, 0.1, 0,0,0,0,0,-0.1,-0.2), xlabx= 4, xlaby=-1.75, tsize=1.5)

```

### Signal-suppression
Uncertainty visualisation is required for transparency. 
The two choropleth maps that appear to be identical in @fig-choropleth highlight the issues with simply electing to ignore uncertainty. 
This sentiment appears frequently in the uncertainty visualisation literature.
Some authors suggest uncertainty is important to include as it communicates the legitimacy (or illegitimacy) of the conclusion drawn from visual inference [@Correll2014; @Kale2018; @Griethe2006]. 
Some authors have said that uncertainty should be included to degree of confidence or trust in the data [@Boukhelifa2012; @Zhao2023]. 
Some authors directly connect uncertainty visualisation to hypothesis testing as it ensures the validity of a statement [@Hullman2020a; @Griethe2006], but allows for a proportional level of trust that is more detailed than the binary results of a hypothesis test [@Correll2014; @Correll2018]. 
Some authors even go so far as to claim that failing to include uncertainty is akin to fraud or lying [@Hullman2020a; @Manski2020].

This consensus leads us to understand that uncertainty visualisation is motivated by the need for a sort of "visual hypothesis test". 
A successful uncertainty visualisation would act as a "statistical hedge" for any inference we make using the graphic. 
Since the purpose of a visualisation is to give a quick gist of the information [@Spiegelhalter2017], this hedging needs to be communicated visually without the need for complicated calculations. 
If we refer to the conclusion we draw from a graphic to be its signal and the variance that makes this signal harder to identify as the "noise", we can summarise the above information into three key requirements. A good uncertainty visualisation needs to:

  1) Reinforce justified signals to encourage confidence in results
  2) Hide signals that are overwhelmed by noise to prevent unjustified conclusions
  3) Perform tasks 1) and 2) in a way that is proportional to the level of confidence in those conclusions.

As @fig-choropleth showed, visualisations that are unconcerned with uncertainty have no issue showing justified signals, but struggle with the display of unjustified signals. 
Therefore, we call this approach to uncertainty visualisation as "signal-suppression" since it primarily differentiates itself from from the normal "noiseless" visualisation approach through criteria (2). 
That is, the main difference between an uncertainty visualisation and a normal visualisation is that an uncertainty visualisation should prevent us from drawing unjustified conclusions.

### Uncertainty as a signal
Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss if we did not mention these alternative approaches. 
Some authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017]. 
Other authors do not describe uncertainty as important for decision making, but rather explicitly state it as a variable of importance in of itself [@Blenkinsop2000]. 
While uncertainty can provide useful information in decision making, it is important to recognise that the "uncertainty" in these cases is not acting as uncertainty at all. 
It is acting as signal.

This is obvious for the cases where we are explicitly interested in the variance or error, as we are literally trying to draw conclusions about an statistic that is used to describe uncertainty. 
The same is true for visualisations made for decision making, but it is less overt. 
This is easiest to understand with an example.

Imagine you are trying to decide if you want to bring an umbrella with you to work. 
An umbrella is annoying to bring with you, so you only want to pack it if the chance of rain is greater than 10%. 
Unfortunately, your weather prediction app only provides you with the predicted daily rainfall. 
Therefore, your decision will be improved with the inclusion of uncertainty. 
This is *not* because uncertainty in general is important for decision making, but because it gives you the tools required to calculate the *actual* statistic you are basing your decision on. 
In this sense, uncertainty is no more special to decision making than weight is special to a body mass index calculation.

This means the uncertainty visualisations that would perform the best in decision making would simply display the uncertainty statistic we are interested in, such as the variance, or probability of an event, using existing visualisation principles. 
This is precisely what we observe in the literature. 
@fig-exceed depicts an exceedance probability map that was designed as an alternative to the choropleth map to improve decision making under uncertainty [@Kuhnert2018; @Lucchesi2021]. 
A keen viewer may notice that the exceedance probability map is actually just a choropleth map, only the statistic being displayed has changed. 
We are not sure it is productive to categorise this visualisation as an uncertainty visualisation.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-exceed
#| fig-cap: "An exceedance probability map that depict the counties of Iowa where each country coloured acording to the probability that the average temperature exceeds 27. This map is a choropleth map where the variable of interest is a probability."
#| fig-subcap: 
#|   - "Low Variance Data"
#|   - "High Variance Data"
#|   - "Exceedance Probability Map Palette"
#| layout-ncol: 3

# quantile
prob_breaks <- seq(-0.1,1.1, length.out=9)
exeed_data <- my_map_data |> 
  as_tibble() |>
  mutate(xprob = 1- pnorm(27, mean=temp, sd=sqrt(variance))) |>
  mutate(xprob = cut(xprob, breaks=prob_breaks, labels=seq(8)))

# Exceed Prob Map
p2a <- exeed_data |>
  filter(variance_class=="lowvar") |>
  ggplot() +
  geom_sf(aes(fill = xprob, 
              geometry = geometry), colour=NA) + 
  scale_fill_manual(values = basecols) +
  theme_void() + 
  theme(legend.position = "none")

p2b <- p2a %+% filter(exeed_data, variance_class=="highvar")

p2a
p2b
show_pal(basecols, ncol=8, borders=NA, myxlab = "P(Temperature>27)", breaks = c(0, prob_breaks[2:8], 1), textnudge = c(0.2, 0.1, 0,0,0,0,0,-0.1,-0.2), xlabx= 4, xlaby=-1.75, tsize=1.2)
```

There seem to be two different definitions of uncertainty visualisation floating around in the literature. 
The first considers *any* visualisation of error, variance, or probability to be an uncertainty visualisation. 
The second believes an uncertainty visualisation is the output of a function that takes a normal visualisation as an input, and transforms it to include uncertainty information. 
The former group believe the purpose of uncertainty visualisation to provide signal about a distribution, while the later believe it should act as noise to obfuscate a signal. 
The lack of explicit distinction between these two motivations leaves the literature muddled and reviewers struggle to understand if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014]. 
This disagreement creates constant contradictions in what the literature considers to be an uncertainty visualisation. 
For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, and @Wickham2011 suggests their product plot framework, which includes histograms and bar charts, should be extended to include uncertainty. 
However, pie charts, bar charts and histograms have all been used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012]. 
If you view an uncertainty visualisation as a function applied to an existing graphic, then you would not see a pie chart or bar chart as uncertainty visualisations.
These charts are they are yet to have the uncertainty visualisation function applied to them. 
If you view an uncertainty visualisation as any graphic that depicts a statistic then there are no limitations on which graphics can or cannot be uncertainty visualisations.

When we use the term uncertainty visualisation to refer to graphics that simply communicate a variance or probability, we are classifying visualisations by the data they display, not their visual features. 
Graphics, just like statistics, are not defined by their input data. 
A scatter plot that compares mean and a scatter plot that compares variances are both scatter plots. 
Given that there is no special class of visualisation for *other* statistics (such the median or maximum) there is no reason to assume visualisations that simply depict a variance, error, or probability to be special. 
Some authors implicitly suggest that visualisations of variance or probability are differentiated due to the psychological heuristics involved in interpreting uncertainty [@Hullman2019].
While it is true that heuristics lead people to avoid uncertainty [@Spiegelhalter2017], there is no evidence that this psychological effect translates to issues with the visual representation of uncertainty. 
Again, given that we do not make these same visual considerations for other variables that elicit distaste or irrational behaviour, there is no reason to assume this is what makes uncertainty visualisation so special.

This leads us to the conclusion that the visualisations made for the purpose of displaying information about uncertainty statistics are not uncertainty visualisations. 
These graphics are just normal information visualisations, and authors can follow existing principles of graphical design. 
We focus on the perspective that uncertainty visualisation serves to obfuscate signal, and an uncertainty visualisation is a variation on an existing graphic that gives it the ability to suppress false signals.

Of course, there is nothing wrong with explicitly visualising variance, error, bias, or any other statistic used to depict uncertainty as a signal. 
Just like any other statistic, these metrics provide important and useful information for analysis and decisions. 
However, there is no interesting visualisation challenge associated with these graphics, and they do not require any special visualisation techniques. 
The uncertainty in these graphics are acting as a signal variable, and they should be treated as such.

>>>>

---
title: "Paper Summary: Uncertainty Visualisation"
author: 
  - name: Harriet Mason
    url: https://harrietmason.netlify.app/
    orcid: 0009-0007-4568-8215
    email:  harriet.m.mason@gmail.com
    affiliation:
    - name: Monash University
      department: Department of Econometrics and Business Statistics
      city: Melbourne
      country: Australia
  - name: Dianne Cook
    url: https://dicook.org
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu
    affiliation:
    - name: Monash University
      department: Department of Econometrics and Business Statistics
      city: Melbourne
      country: Australia
  - name: Sarah Goodwin
    url: https://www.linkedin.com/in/smgoodwin/
    orcid: 0000-0001-8894-8282
    email: Sarah.Goodwin@monash.edu
    affiliation:
    - name: Monash University
      department: Department of Human Centred Computing
      city: Melbourne
      country: Australia
  - name: Emi Tanaka
    url: https://emitanaka.org/
    orcid: 0000-0002-1455-259X
    email: Emi.Tanaka@anu.edu.au
    affiliation: 
    - name: The Australian National University
      department: Biological Data Science Institute
      city: Canberra
      country: Australia
  - name: Susan VanderPlas
    url: https://srvanderplas.github.io
    orcid: 0000-0002-3803-0972
    email: susan.vanderplas@unl.edu
    affiliation:
    - name: University of Nebraska–Lincoln
      department: Statistics Department
      city:  Lincoln
      country: United States
bibliography: summary.bib
nocite: |
  @*
abstract: Uncertainty visualisation reviews often take the definition and purpose of an uncertainty visualisation to be self evident which results in a large amout of conflicting information. This conflict largely stems from a conflation between uncertainty visualisations designed for to visualise uncertainty as signal and those used to visualise uncertainty as noise. We coin the term "signal suppression" to describe a visualisation that is designed for preventing false conclusions, as the approach demands that the signal (i.e. the collective take away of the estimates) is suppressed by the noise (i.e. the variance on those estimates). We argue that the current standards in visualisation suggest that uncertainty visualisations designed to extract uncertain information should not be considered uncertainty visualisations at all. Therefore, future work should focus on signal suppression. Effective signal suppression requires us to communicate the signal and the noise as a single "validity of signal" variable, and doing so proves to be difficult with current methods. We illustrate current approaches to uncertainty visualisation by showing how they would change the visual apprearance of a choropleth map. These maps allow us to see why some methods succeed at signal suppression, while others fall short. Evaluating visualisations on how well they perform signal suppression also proves to be difficult, as it involves measuring the effect of noise, a variable we typically try to ignore. We suggest authors use qualitative studies or compare uncertainty visualisations to the relevant hypothesis tests. 
date: last-modified
toc: false
number-sections: true
latex-clean: true
format:
  jasa-pdf:
    keep-tex: true  
    journal:
      blinded: false
  jasa-html: default
fig-valign: bottom
cap-location: bottom
editor_options: 
  chunk_output_type: console
---
